A novel approach to chemical tagging in lower-dimensional latent space

### 2019-01-18 article/mcfa.tex
exp2 + 1 figs

\documentclass[twocolumn]{aastex62}
\usepackage{bm}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{comment}

@@

We generated a data set with ${\NumData = 1}$00,000 data points, each with
$\NumDimensions = 15$ dimensions, and assumed that those data are generated by
latent space of $\NumLatentFactors = 5$ ($\NumLatentFactors \times \NumDimensions$) factor loads, and there are $\NumComponents = 20$
clusters in the latent space. The relative weights $\vec\weight$
are drawn from a multinomial distribution and the means of the clusters
in factor scores $\scoremeans$ are drawn from a standard normal
distribution. The off-diagonal entries in the covariance matrices in factor scores $\scorecovs$ are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$. The variance in 
each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads, and adding variance.

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 data generated as part of our toy model. The top 
		 	 panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
			 and the lower panel shows
			 the pseudo-BIC
			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
			 lowest value in each panel.}
    \label{fig:experiment-1-gridsearch}
\end{figure}

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, the
Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
\end{equation} 
\noindent{}where $Q$ is the number of model parameters:
\begin{equation}
	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
\end{equation}
%\todo{\noindent{}and the Message Length.}

These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the log likelihood increases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value is found at $\NumLatentFactors = 5$
and $\NumComponents = 21$, close to the true values ($\NumLatentFactors_\textrm{true} = 5$,
$\NumComponents_\textrm{true} = 20$).

We show a corner plot of the generated data in Figure~\ref{fig:exp1-data},
demonstrating the clustering (colour) and structure varying structure in data space.
Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
and allowing clustering in the latent space provides a better description of the data.
Adding components to the model improves the log-likelihood, even with a single latent factor,
but the addition of just one latent factor improvers the log-likelihood more than adding
20 clusters. Not much can be said for this example because the true data generating process 
is drawn from a higher number of latent factors, but this does illustrate how 
clustering in high dimensional data can be better described by latent factors with 
clustering in the lower dimensional latent space.

Given our estimate of the number of latent factors and components ($\NumComponents = 21$, $\NumLatentFactors = 5$), in Figure~\ref{fig:exp1-latent} we show the estimated latent
space (factor scores $\factorscores$) where each point is coloured by it's inferred
associated component. After applying an orthogonal rotation (see following section)
to the estimated latent space, we show the estimated factor scores (Figure~\ref{fig:exp1-factor-scores-comparison}) and estimated factor loads (Figure~\ref{fig:exp1-factor-loads-comparison})
compared to the true values. 
Similarly, in Figure~\ref{fig:exp1-specific-variance} we show the estimated specific scatter
in each data dimension compared to the true value. 
\todo{The agreement is excellent in all model parameters.}

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
    \caption{The inferred latent scores $\factorscores$ for Experiment~1. 
             Points are coloured by the inferred associated component.}
    \label{fig:exp1-latent}
\end{figure*}

%\begin{figure*}
%	\includegraphics[width=1.0\textwidth]{experiments/toy-factors-init-randomly.png}
%	\caption{Latent factor loads inferred for models with increasing numbers
%			 of latent factors and increasing numbers of cluster components
%			 in Experiment~1, where the latent factors and component assignments
%			 were initialised randomly.
%				The top left panel shows the results of the model with
%			 $\NumLatentFactors = 1$ and $\NumComponents = 1$. 
%			 The latent factors have each been rotated by a
%			 rotation matrix such 
%			 that the estimated latent factors in each model are ordered 
%			 and oriented to be as similar as the latent factors estimated from
%			 the model with only one latent factor, up to orthogonality such
%			 that $\vec{R}\vec{R}\transpose = \eye$.
%			 }
%	\label{fig:toy-factors-init-randomly}
%\end{figure*}

%\todo{This structure goes away and becomes more consistent if we change the initialisation procedure}

%\todo{figure of toy factors init svd kmeans++}

%\todo{log-likelihood of different initialisation routines}

\begin{comment}
The results of this experiment confirm that the model is indeterminate, in 
that there are multiple equally valid solutions that can exactly reproduce
the data. It also highlights the importance of initialisation, demonstrating
how the initialisation of the common latent factors can impact the final
model parameters. These two points are related, as one could initialise the
latent factors randomly (or from noise) and multiply that model with a
rotation matrix $\mathbf{R}$ that rotates the latent space such that the
common factor loads are aligned closer to some prior expectation that we have
for the common factor loads (e.g., in that they should represent plausible
astrophysical yields). Motivated by these results and indeterminate nature
of the model, in Experiment~2 we initialise the common factor loads
to represent known astrophysical processes such that they can be interpreted
in context with the data.

\subsection{Experiment 2: A minimal set of nucleosynthetic trace elements in \Galah}

In this experiment we further investigate the effects of initialising the model parameters.
Here we make use of a minimal set of nucleosynthetic trace elements
using the second \Galah\ data release \citep{Buder:2018}. Specifically, we use
reliable chemical abundance measurements of Na (an odd-Z element), Mg and Ca (alpha-elements),
Ni, Fe, and Cr (iron-peak elements), Ba and Y (s-process tracers), and Eu (an r-process tracer).
We assume that there are four latent factors ($J = 4$) and four components ($K = 4$),
and we initialise the latent factors and component parameters in different combinations.

To initialise the latent factors using astrophysical insight (or domain expertise)
we make the following steps:
\begin{enumerate}
	\item We initialise all entries of $\factorloads$ to be drawn from $\mathcal{N}\left(0, 10^{-2}\right)$.
	\item For the first latent factor we set the entries that correspond to Ni, Fe, and Cr to be $+1$ (e.g., Fe-peak).
	\item For the second latent factor we set the entries that correspond to Mg and Ca to be $+1$ (e.g., $\alpha$-elements).
	\item For the third latent factor we set the entries that correspond to Ba and Y to be $+1$ (e.g., the s-process).
	\item For the fourth latent factor we set the entry that corresponds to Eu to be $+1$ (e.g., the r-process).
\end{enumerate}

%The latent factors can be initialised in four different ways: (1) with entries drawn randomly from $\mathcal{U}\left(-1, +1\right)$, (2) with noisey draws $\mathcal{N}\left(0, 10^{-2}\right)$, (3) using singular value decomposition, and (4) using astrophysical insight to initialise the factor loads.

The component parameters are either initialised randomly, or using \texttt{K-means++} algorithm in
the latent space given the initial estimate of the latent factors. For each combination of 
initialisation procedure, we run 30 optimisations and record the log likelihood once the E-M 
procedure has reached a threshold of $10^{-8}$.
The results of this experiment are shown in Figure~\ref{fig:experiment-4-log-L}.
In most cases the final log likelihood is within the threshold of $10^{-8}$ between
different optimisation runs of the same initialisation procedure. However, it is clear
that at least some initialisations result in a local optimum, emphasising the importance in
optimising from many initialisations.

Figure~\ref{fig:experiment-4-log-L} demonstrates that even though the initialisation procedure
can impact the final estimate of the model parameters, the resulting log likelihood values
are within the convergence threshold. This is, again, a consequence of the indeterminate
nature of the model specification. However, the advantage here is that these results
demonstrate that we \emph{can} initialise the latent factors with domain knowledge,
without a significant detriment to the final log likelihood. Similarly, in Figure~\ref{fig:experiment-4-astrophysical-latent-factors}
we show that the initial latent factors are indeed different from the final latent factors,
and that there is some variance in our estimates of the latent factors from 30 initialisations.
\todo{something to say here about not biasing our results?}

The results from this experiment demonstrate that we can initialise the latent factors
using domain knowledge without incurring strong biases in our final estimates of the
latent factors, or introducing a significant detriment to the log likelihood, as long
as we initialise the model many (e.g., $\gtrsim 5$) times.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{experiments/galah-experiment-4-log-L-hist.png}
	\caption{Log likelihoods $\mathcal{L}\left(\data|\vec\Psi\right)$ of optimised
			 models using \Galah\ data in Experiment 4 with four latent factors
			 and four components ($J = 4$, $K = 4$). The columns indicate different
			 methods to initialise the component parameters ($\pi$, $\scoremeans$,
			 $\scorecovs$), and the rows represent different methods to initialise
			 the latent factors ($\factorloads$). Each combination of initialisation
			 procedure has 30 models where E-M was run until a threshold of $10^{-8}$
			 was reached in $\log{L}$.}
	\label{fig:galah-experiment-4-log-L}
\end{figure}

\todo{experiment-4-astrophysical-latent-factors}
\end{comment}

@@

In this experiment we use detailed chemical abundances of metal-poor stars 
\citep{Barklem:2005} to demonstrate the utility and limitations
in interpreting the latent factors as nucleosynthetic processes. The data 
can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), that trace multiple nucleosynthetic pathways. Specifically the chemical
abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak), as well as Sr, Y, Ba, and Eu ($s$/$r$-process).

We performed a grid search for the optimal number of latent factors $\NumLatentFactors$ and 
components $\NumComponents$, starting from $\NumLatentFactors = 1$ and $K = 1$, and trialling up until $\NumLatentFactors = \todo{8}$ and $\NumComponents = 5$. We initialised
the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
and BIC recorded for the optimised set of model parameters for each permutation 
of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
five latent factors are preferred, with no clustering.

Some technical discussion is warranted before we interpret the estimated factor loads.
In principle the latent factors are analogous to nucleosynthetic yields, but 
these can only be identified up to an orthogonal rotation. That is to say that
if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
then our estimates of those latent factors $\factorloads_\textrm{est}$ do not need
to be identical to the true values. For example, the ordering of the estimated factors
could be different from the true factors, and the ordering of the dimensionality
in latent space would be correspondingly different. Since the ordering of the 
factor loads is not prescribed during expectation-maximization, there is no 
assurance (or requirement) that our factor loads match the true factor loads.

Another possibility is that the estimated factor loads could be flipped in sign 
relative to the true factor loads, and the scores would similarly be flipped. 
In both of these situations (reordering or flipped signs) the log likelihood 
given the data and the estimated factor loads $\factorloads_\textrm{est}$ would 
be identical to the log likelihood given the data and the true factor loads 
$\factorloads_\textrm{true}$
despite the difference in ordering and sign. These examples serve to illustrate a more 
general property that the factor loads and factor scores can be orthogonally 
rotated by \emph{any rotation matrix}\footnote{A rotation matrix is valid if 
$\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
$\factorloads_\textrm{est}$ could therefore appear very different from the true 
values, but they only differ by an orthogonal rotation.

If the true values are known then it is trivial to find a rotation matrix $\vec{R}$
that rotates the estimated factor loads to be as close to the true factor loads as
possible. However, when
the true values are not known, this property of factor analysis limits the
extent to which we can interpret the latent variables in our model. For example,
if we seek to interpret the factor loads as being something analagous to 
nucleosynthetic processes, then the estimated factor loads will need to be 
rotated to something that looks astrophysically plausible before we could 
attribute one latent factor as being a particular nucleosynthetic process. 
Although having to perform this rotation is not comforting, there is utility in
trying to identify the latent factors because the factors can only be rotated by
some valid rotation matrix. That is to say that we will not only exactly what
we have put in: the rotated factors will simply be the closest set of factor
loads that are consistent with the data, allowing us to identify the factor
loads that are astrophysically plausible.

Using our preferred model with $J = 5$ and $K = 1$, we rotated the estimated
factor loads in order to be astrophysically plausible. Specifically we constructed
a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
all entries were zero except for the following entries:
\begin{itemize}
	\item the entry corresponding to Al in the first latent factor to represent
		  odd-Z element production,
	\item the entries corresponding to Ca and Mg in the second latent factor to
		  represent $\alpha$ particle production,
	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
		  latent factor to represent the Fe-peak group,
	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
		  to represent products of the slow neutron capture process, and
	\item the entries corresponding to Eu in the fifth latent factor to represent
		  products of the rapid nucleosynthesis process.
\end{itemize}
These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
orthonormal. We optimise the $J$ angles needed to make the estimated factor loads
$\factorloads_\textrm{est}$ as close to the astrophysical factor loads $\factorloads_\textrm{A}$.
In practice we take the product of many  Givens rotation matrices \citep{Givens},
which reduces to Euler angle rotation in three or fewer dimensions.
%\footnote{Clifford
%algebra offers a more efficient algorithm that is more suitable for high dimensions, 
%but we found Givens rotation matrices to be suitable for our purposes here.}
In Figure
\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
preferred model, where the shaded regions indicate the projected uncertainty on
the rotation angles -- not including any uncertainty on the factor loads themselves.

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.pdf}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 \citet{Barklem:2005} data in Experiment~2.
    		 The top panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
			 The cross at
			 $J=8$ and $K=5$ indicates that the model could not converge without
			 adding a regularisation constant to the diagonals of $\scorecovs$.
			 The marker indicates the lowest value in each panel, showing that
			 $J = 5$ and $K = 1$ is preferred by BIC.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

@@

>>>>

\documentclass[twocolumn]{aastex62}
\usepackage{bm}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{comment}
%\usepackage{layouts}

@@

We generated a data set with ${\NumData = 1}$00,000 data points, each with
$\NumDimensions = 15$ dimensions, and assumed that those data are generated by
latent space of $\NumLatentFactors = 5$ ($\NumLatentFactors \times \NumDimensions$) factor loads, and there are $\NumComponents = 20$
clusters in the latent space. The relative weights $\vec\weight$
are drawn from a multinomial distribution and the means of the clusters
in factor scores $\scoremeans$ are drawn from a standard normal
distribution. The off-diagonal entries in the covariance matrices in factor scores $\scorecovs$ are drawn from a gamma distribution $\scorecovs_{\numcomponents,i,i} \sim \vec\Gamma\left(1\right)$. The variance in 
each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
We randomly draw a  $\NumDimensions \times \NumDimensions$ matrix from a Haar distribution \citep{Haar:1933},
which is uniform on the special orthogonal group $\textrm{SO}(n)$ and therefore guaranteed to return an orthogonal
matrix with a determinant of unity \citep{Stewart:1980}.
We denote the $\NumLatentFactors \times \NumDimensions$ left-most region of this
matrix to be $\mathbf{H}$, and by taking $\factorloads_\ast$ to be the Cholesky decomposition of $\mathbf{H}\transpose \mathbf{H}$, we find the factor loads
\begin{eqnarray}
 	\factorloads = \mathbf{H}\left(\left(\factorloads_\ast\right)^{-1}\eye\right)
\end{eqnarray}
\noindent{}which ensures that the true factor loads is a $\NumLatentFactors \times \NumDimensions$
matrix of mutually orthogonal vectors and that $\factorloads\transpose \factorloads = \eye$.
The $\numdata$th data point (which belongs to the $\numcomponents$th cluster) is then
generated by drawing $\factorscores_{\numdata} \sim \mathcal{N}(\scoremeans_\numcomponents,\scorecovs_\numcomponents)$, projecting by the factor loads $\factorloads$, and adding variance $\specificvariance$.

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 data generated as part of our toy model. The top 
		 	 panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$ and the bottom panel shows the BIC (Equation \ref{eq:bic}).
			 The white marker indicates the 
			 lowest value in each panel, which is joined by a black line from the true value.}
    \label{fig:experiment-1-gridsearch}
\end{figure}

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
For each $\NumLatentFactors$, $\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, the
Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
\end{equation} 
\noindent{}where $Q$ is the number of model parameters:
\begin{equation}
	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
\end{equation}

These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the log likelihood increases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value is found at $\NumLatentFactors = 5$
and $\NumComponents = 20$, identical to the true values.

We show a corner plot of the generated data in Figure~\ref{fig:exp1-data},
demonstrating the clustering (colour) and structure varying structure in data space.
Despite this clustering in data space, it is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
and allowing clustering in the latent space provides a better description of the data.
Adding components to the model improves the log-likelihood, even with a single latent factor,
but the addition of just one latent factor improvers the log-likelihood more than adding
20 clusters. Not much can be said for this example because the true data generating process 
is drawn from a higher number of latent factors, but this does illustrate how 
clustering in high dimensional data can be better described by latent factors with 
clustering in the lower dimensional latent space.
The inferred clustering in latent space is shown in Figure~\ref{fig:exp1-latent},
where each point is coloured by the inferred associated component.

Some technical discussion is warranted before we compare our estimated model
parameters to the true values. The latent factors in this model are only
identifiable up to an orthogonal rotation. That is to say that
if the data were truely generated by latent factors $\factorloads_\textrm{true}$,
then our estimates of those latent factors $\factorloads_\textrm{est}$ do not need
to be identical to the true values. For example, the ordering of the estimated factors
could be different from the true factors, and the ordering of the dimensionality
in latent space would be correspondingly different. Since no constraint is
placed on the ordering of the factor loads during expectation-maximization,
there is no assurance (or requirement) that our factor loads match the true factor loads.

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp1-latent.pdf}
    \caption{The inferred latent scores $\factorscores$ for Experiment~1
    		 from a model with $\NumLatentFactors = 5$ latent factors and
		 	 $\NumComponents = 20$ multivariate normal components in
			 latent space.
             Points are coloured by the inferred component.}
    \label{fig:exp1-latent}
\end{figure*}

Another possibility is that the estimated factor loads could be flipped in sign 
relative to the true factor loads, and the scores would similarly be flipped. 
In both of these situations (reordering or flipped signs) the log likelihood 
given the data and the estimated factor loads $\factorloads_\textrm{est}$ 
%\citep[or Kullback-Leibler divergence;][]{Kullback:1951}
would be identical to the log likelihood given the data and the true factor loads 
$\factorloads_\textrm{true}$
despite the difference in ordering and sign. These examples serve to illustrate a more 
general property that the factor loads and factor scores can be orthogonally 
rotated by \emph{any valid rotation matrix}\footnote{A rotation matrix is valid if 
$\vec{R}\vec{R}\transpose = \vec{I}$} $\vec{R}$. The estimated factor loads 
$\factorloads_\textrm{est}$ could therefore appear very different from the true 
values, but they only differ by an orthogonal rotation.

% TODO: Make single figure with bigger text labels.
\begin{figure*}
	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-loads.pdf}
	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-scores.pdf}
	\includegraphics[width=0.33\textwidth]{experiments/exp1-compare-specific-scatter.pdf}
    \caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~1. The agreement is excellent.}
    \label{fig:exp1-compare}
\end{figure*}

We took the model with the preferred number of latent factors and components found
from a grid search ($\NumComponents = 20$, $\NumLatentFactors = 5$; which are also
the true values) and applied an orthogonal rotation to the latent space to be as
close as possible to the true values. The rotation matrix $\mathbf{R}$ was found
by solving for $\NumLatentFactors$ unknown angle parameters, each of which is used
to construct a Givens rotation matrix, and then we take the product of those Givens
matrices. This process reduces to Euler angle rotation in three or fewer dimensions.
This process rotates the latent space
($\factorloads$, $\scoremeans$, $\scorecovs$), but has no effect on the model's 
predictive power: the log-likelihood \citep[or the Kullback-Leibler divergence;][]{Kullback:1951} under the
rotated model is indistinguishable from the unrotated model.
In Figure~\ref{fig:exp1-compare} we show the estimated factor loads $\factorloads$,
factor scores $\factorscores$, and specific variances $\specificvariance$ compared
to the true values. The agreement is excellent in all model parameters.

@@

In this experiment we use detailed chemical abundances of metal-poor stars 
\citep{Barklem:2005} to demonstrate the utility and limitations
in interpreting the latent factors as nucleosynthetic processes. The data 
can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), and those 14 chemical elements trace multiple nucleosynthetic pathways. Specifically the chemical
abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process).

We performed a grid search for the optimal number of latent factors 
$\NumLatentFactors$ and components $\NumComponents$, starting from 
$\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
$\NumLatentFactors = \todo{8}$ and $\NumComponents = 5$. 
We found that models with higher numbers of latent factors were unable to
converge without adding a regularization constant to the diagonals of the
covariance matrices $\scorecovs$ in latent space. Although it is common
to add a regularization term to the diagonals of covariance matrices
when performing expectation-maximization, we found that our existing
grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
and $\NumComponents$, and chose not to add any regularization term.

We initialised
the model parameters randomly for each permutation of $J$ and $K$, and continued
the expectation-maximization algorithm
until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
and BIC recorded for the optimised set of model parameters for each permutation 
of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
five latent factors are preferred, with no clustering.

Using our preferred model with $J = 5$ and $K = 1$, we sought to find a
rotation matrix that would allow us to identify plausible astrophysical
processes in the factor loads.
Although having to perform this rotation is not comforting, there is utility in
trying to identify the latent factors because the factors can only be rotated by
some valid rotation matrix. That is to say that we will not only exactly what
we have put in: the rotated factors will simply be the closest set of factor
loads that are consistent with the data, allowing us to identify the factor
loads that are astrophysically plausible. We constructed
a $J \times D$ astrophysical factor load matrix $\factorloads_\textrm{A}$ where
all entries were zero except for the following entries:
\begin{itemize}
	\item the entry corresponding to Al in the first latent factor to represent
		  odd-Z element production,
	\item the entries corresponding to Ca and Mg in the second latent factor to
		  represent $\alpha$ particle production,
	\item the entries corresponding to Ni, Co, Fe, Mn, Cr, Ti, and Sc in the third
		  latent factor to represent the Fe-peak group,
	\item the entries corresponding to Sr, Y, and Ba in the fourth latent factor
		  to represent products of the slow neutron capture process, and
	\item the entries corresponding to Eu in the fifth latent factor to represent
		  products of the rapid nucleosynthesis process.
\end{itemize}
These entries were set to $1/\sqrt{E}$ where $E$ is the number of non-zero entries
in that latent factor, in order to ensure that $\factorloads_\textrm{A}$ is
mutually orthogonal. We then found the valid rotation matrix $\vec{R}$ that most
closely approximated the astrophysical loads $\factorloads_\textrm{A}$.
In Figure~\ref{fig:exp2-factor-loads} we show the rotated factor loads found by our
preferred model, where the shaded regions indicate \emph{only} the projected uncertainty on
the rotation angles -- and does not include any uncertainty on the factor loads 
themselves.

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.pdf}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 \citet{Barklem:2005} data in Experiment~2.
    		 The top panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $\NumLatentFactors$ and number 
			 of clusters $\NumComponents$, and the lower panel shows the BIC (Eq.~\ref{eq:bic}).
			 %The cross at
			 %$\NumLatentFactors=8$ and $\NumComponents=5$ indicates that the model could not converge without
			 %adding a regularisation constant to the diagonals of $\scorecovs$.
			 The marker indicates the lowest value in each panel, showing that
			 $\NumLatentFactors = 5$ and $\NumComponents = 1$ is preferred by BIC.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{experiments/exp2-latent-contributions.pdf}
	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. \todo{elaborate latent factors}}
    \label{fig:exp2-latent-contribution}
\end{figure}
